Shallow EDSLs and Object-Oriented Programming

### 2016-06-20 paper/paper.tex
Separate transformations from S2; Discuss sophisticated optimizations
Draft Section 4
Short related work
More text

\documentclass[10pt,preprint,nocopyrightspace]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{color}
\usepackage{listings}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{theorem}
\usepackage{bbm}
\usepackage{pgf}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{mdwlist}
\usepackage{morefloats}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{mathptmx}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
\usepackage{tablefootnote}
\usepackage{enumitem}
\usepackage{stmaryrd}

\newcommand{\name}{\textbf{@Family}\xspace}
\newcommand{\dsl}{\emph{Scans}\xspace}
\newcommand{\interp}{\textsc{Interpreter}\xspace}

@@

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a DSL into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). Given
that OO languages have evolved over more than 50 years
to improve the use of procedural abstraction, they ought to have some
advantages to encode shallow EDSLs.

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
EDSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Scala. From the \emph{semantic}
and \emph{modularity} point of view the Scala version has clear advantages 
over the Haskell version.

@@

%===============================================================================
\input{sections/Introduction.tex}
\input{sections/ShallowOO.tex}
\input{sections/Overview.tex}
\input{sections/Tool.tex}
\input{sections/CaseStudy.tex}
\input{sections/RelatedWork.tex}
\input{sections/Discussion.tex}
%\acks

>>>>

\documentclass[10pt,preprint,numbers,nocopyrightspace]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{color}
\usepackage{listings}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{theorem}
\usepackage{bbm}
\usepackage{pgf}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{mdwlist}
\usepackage{morefloats}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{mathptmx}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
\usepackage{tablefootnote}
\usepackage{enumitem}
\usepackage{stmaryrd}
\usepackage{syntax}

\newcommand{\name}{\textbf{@Family}\xspace}
\newcommand{\dsl}{\emph{scans}\xspace}
\newcommand{\interp}{\textsc{Interpreter}\xspace}

@@

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a DSL into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). 
%Given
%that OO languages have evolved over more than 50 years
%to improve the use of procedural abstraction, they ought to have some
%advantages to encode shallow EDSLs.
This paper argues that OOP abstractions 
(including \emph{inheritance}, \emph{subtyping} and
\emph{type-refinement}) 
increase the modularity and reuse of shallow
EDSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Scala. From the \emph{semantic}
and \emph{modularity} point of view the Scala version has clear advantages 
over the Haskell version.

@@

%===============================================================================
\input{sections/Introduction.tex}
\input{sections/ShallowOO.tex}
\input{sections/Transformations.tex}
\input{sections/Modularity.tex}
\input{sections/RelatedWorkShort.tex}
%\input{sections/Overview.tex}
%\input{sections/Tool.tex}
%\input{sections/CaseStudy.tex}
%\input{sections/RelatedWork.tex}
%\input{sections/Discussion.tex}
%\acks

### 2016-06-20 paper/sections/Introduction.tex
More text

Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. Since the {\sc Composite}
pattern is normally accepted to be a way to encode ASTs, it would be
reasonable to say that \emph{according to definition of deep embedding
  above, the OO approach classifies as a deep
  embedding}. However, as we shall argure in the remainder of the
paper, another possible interpretation is that the OO approach is
really a shallow embedding. At least some authors~\cite{} seem to agree with 
the latter interpretation, but we believe that the current definition
leaves some space open to interpretation.

@@

To avoid confusion and make the terminology more precise, we first 
propose distinguishing EDSLs in terms of the data
abstraction used to model the language constructs instead.  We follow
Reynold's classification~\cite{} of data abstractions:
\emph{procedural abstraction} and \emph{user-defined types}. It is
clear that shallow embeddings use \emph{procedural
  abstraction}~\cite{}: the DSLs are modelled by interpretation
functions. Therefore, the other implementation option for EDSLs is to use
\emph{user-defined types}. In Reynolds terminology user-defined types
mean disjoint union types, which are nowadays commonly available in
modern languages as \emph{algebraic datatypes}. Disjoint union types 
can also be emulated in OOP using the {\sc Visitor} pattern. A
distinction based on data abstraction provides a remedy for possible
misinterpretation. An EDSL implemented with algebraic
datatypes falls into the category of user-defined types (deep embedding), 
while a Composite-based OO implementation falls under procedural
abstraction (shallow embedding). For the rest of the paper we identify shallow EDSLs 
with EDSLs implemented using procedural abstraction.

This paper shows that OOP languages have
advantages for the implementation of shallow embeddings.
As Cook~\cite{} argued procedural abstraction is the essence of
Object-Oriented Programming. If we accept Cook's view, the
implementation of a shallow DSL in OOP languages should simply
correspond to a standard object-oriented program. Given
that Object-Oriented Languages have evolved over more than 50 years 
to improve the use of procedural abstraction, they ought to have some 
advantages to encode shallow EDSLs.

We show that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.

This paper has two goals:

@@

\item To argue that

>>>>

Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. Since the {\sc Composite} or {\sc Interpreter}
patterns are normally accepted to provide a way to encode ASTs, it would be
reasonable to say that \emph{according to definition of deep embedding
  above, the OO approach classifies as a deep
  embedding}. However, as we shall argure in the remainder of the
paper, another possible interpretation is that the OO approach is
really a shallow embedding. At least some authors~\cite{} seem to
implicitly agree with 
the latter interpretation.

@@

To avoid confusion and make the terminology more precise,
this paper distinguishes EDSLs in terms of the data
abstraction used to model the language constructs instead.  We follow
Reynold's classification~\cite{} of data abstractions:
\emph{procedural abstraction} and \emph{user-defined types}. It is
clear that shallow embeddings use \emph{procedural
  abstraction}~\cite{}: the DSLs are modelled by interpretation
functions. Therefore, the other implementation option for EDSLs is to use
\emph{user-defined types}. In Reynolds terminology user-defined types
mean disjoint union types, which are nowadays commonly available in
modern languages as \emph{algebraic datatypes}. Disjoint union types 
can also be emulated in OOP using the {\sc Visitor} pattern. A
distinction based on data abstraction provides a remedy for possible
misinterpretation. An EDSL implemented with algebraic
datatypes falls into the category of user-defined types (deep embedding), 
while a Composite-based OO implementation falls under procedural
abstraction (shallow embedding). For the rest of the paper we identify shallow EDSLs 
with EDSLs implemented using procedural abstraction.

As Cook~\cite{} argued
procedural abstraction is the essence of Object-Oriented
Programming. If we accept Cook's view, the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program. Given that OOP languages have
evolved over more than 50 years to improve the use of procedural
abstraction, they ought to have some advantages to encode shallow
EDSLs.

The main goal of this paper is to show that OOP languages have advantages for the
implementation of shallow embeddings.  
In order to understand how OOP can help with shallow embeddings, lets
first review some of the limitations of shallow embeddings commonly found in
the literature:

\begin{enumerate}

\item {\bf Single Interpretation} An often stated limitation of
  shallow embeddings is that they only support a single
  interpretation.

\item {\bf No Transformations} Another commonly stated limitation 
of shallow embeddings is that they do not support transformations,
preventing optimizations and other useful transformations.

\end{enumerate}

\noindent Both limitations are often used as motivators to switch to deep embeddings.

We show that OOP abstractions, including \emph{(multiple)
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{}, which provides a simple
solution to the \emph{Expression Problem}\cite{} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu,
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. Although from the
\emph{syntactical} point of view there are minor inconveniences in
the Scala version, from the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages.

In summary, our contributions are:

@@

\item {\bf Multiple Modular Interpretations for Shallow Embeddings:} 
  We show that with standard OOP mechanisms it is easy to support multiple modular
  interpretations for shallow embeddings.

\item {\bf Transformations for Shallow Embeddings:} We show that
  transformations are encodable with recursive objects. Moreover, this technique
  can be ported back into functional programming as well.
